---
title: "Loading Seurat Object and Some Examples"
author: "Theresa Alexander"
date: "2/24/2022"
output: html_document
---

# Load the Seurat Object

```{r}
load("seurat_juntti_v2242022.rda")
```

This seurat object comes preloaded with a few pieces of metadata that are important.
<br>
All metadata can be found if you print the object, juntti_OE_new. For example:

```{r}
head(juntti_OE_new)
```

The pieces of metadata that I've added which are important are the following columns in that dataframe:
1. OR_counts: the number of ORs in each cell. Only the cells in the neuronal clusters (including the immature ones) have a value for this. <br>
2. souporcell_assignment: the souporcell cluster which each cell belongs to. These are values between 0 and 3. <br>
3. souporcell_status: This column indicates if a cell is a doublet, singlet, or unassigned. 

Note: This dataset has not been filtered for the percent mt, ncounts, or nfeatures. So to do this, you should do the following beofre doing more downstream analysis:

```{r}
juntti_OE_new <- subset(juntti_OE_new, subset = nFeature_RNA > 200 & nFeature_RNA < 6000 & percent.mt < 5)
```


Now you can do things like run PCA, UMAP, and then plot based on features like the souporcell cluster, status, or other things.

An example of how you would make this UMAP plot (once you do the PCA/UMAP reductions), would be like this:

```{r}
DimPlot(juntti_OE_new, reduction = "umap", group.by = "souporcell_assignment")
```

You can change the "group.by" to be equal to any of the features we've computed (which are the columns you see when you do the head(juntti_OE_new) line of code above).

