---
title: "OR Interrogation"
author: "Theresa Alexander"
date: "3/4/2022"
output: html_document
---

We have a few questions we want to look in to regarding the OR expressing cells.

Tasks from 3/2 meeting: <br>

- what are the most represented ORs we have <br>
- how much fos is a single class expressing (where a class is a single receptor) <br>


# Libraries

```{r, message = FALSE}
library("xlsx")
library("Seurat")
library("dplyr")
library("ggplot2")
library("knitr")
```
# Helper Functions

```{r}
genes_avg_expression <- function(seur_obj, features, slot = "counts"){
  #get counts
  counts <- as.data.frame(GetAssayData(object = seur_obj, slot = slot))
  
  #subset out features OI
  for (i in seq_along(features)){
  counts_oi <- counts %>%
    filter(row.names(counts) %in% features[[i]])

    mean_exp <- colMeans(counts_oi)
    colname <- paste0( names(features)[i], "_Means")
    #mean_exprs[[colname]] <- mean_exp
    seur_obj <- AddMetaData(object = seur_obj, metadata = mean_exp, col.name = colname)

    sum_exp <- colSums(counts_oi)
    colname <- paste0( names(features)[i], "_Sums")
    seur_obj <- AddMetaData(object = seur_obj, metadata = sum_exp, col.name = colname)
  }
  return(seur_obj)
}
```


# Load the Seurat Object

```{r}
load("seurat_juntti_v2242022.rda")
```


# Filtering and normalization

We’ve identified clusters 0 and 3 as being exposed to our pheromone; clusters 1&2 were exposed to the control odor.

## Adding in Fish assignments to metadata

```{r}
fish_assignments <- c(rep("NA", nrow(juntti_OE_new@meta.data)))
fish_assignments[which(juntti_OE_new@meta.data$souporcell_assignment %in% c(0,3))] <- "pheromone_exposed"
fish_assignments[which(juntti_OE_new@meta.data$souporcell_assignment %in% c(1,2))] <- "control"

juntti_OE_new <- AddMetaData(juntti_OE_new, metadata = fish_assignments, col.name = "fish_assignments")
```


```{r}
all.genes <- rownames(juntti_OE_new)
juntti_OE_new <- subset(juntti_OE_new, subset = nFeature_RNA > 200 & nFeature_RNA < 6000 & percent.mt < 5)
juntti_OE_new <- subset(juntti_OE_new, subset = souporcell_status == "singlet")

juntti_OE_new <- NormalizeData(juntti_OE_new, normalization.method = "LogNormalize", scale.factor = 10000)
juntti_OE_new <- ScaleData(juntti_OE_new, features = all.genes)

juntti_OE_new <- FindVariableFeatures(juntti_OE_new, selection.method = "vst", nfeatures = 2000)
juntti_OE_new <- RunPCA(juntti_OE_new, features = VariableFeatures(object = juntti_OE_new))

juntti_OE_new <- FindNeighbors(juntti_OE_new, dims = 1:13)
juntti_OE_new <- FindClusters(juntti_OE_new, resolution = c(0.5))
juntti_OE_new <- RunUMAP(juntti_OE_new, dims = 1:13)
```


## List of OR Genes of interest

```{r}
V1R_genes <- read.xlsx2("~/scratch/reesyxan/juntti_cichlid/data/CR_genes_ABurtoni.xlsx", sheetIndex = "V1R_genes")
OR_genes <- read.xlsx2("~/scratch/reesyxan/juntti_cichlid/data/CR_genes_ABurtoni.xlsx", sheetIndex = "OR_genes")
V2R_genes <- read.xlsx2("~/scratch/reesyxan/juntti_cichlid/data/CR_genes_ABurtoni.xlsx", sheetIndex = "V2R_genes")
TAAR_genes <- read.xlsx2("~/scratch/reesyxan/juntti_cichlid/data/CR_genes_ABurtoni.xlsx", sheetIndex = "TAAR_genes")
```

# How many cells per receptor do we have?

```{r}
juntti_OE_neuronal <- subset(x = juntti_OE_new, idents = c(0,1))

#subset `juntti_OE_neuronal` for only the CR genes
juntti_OE_neuronal_CRs <- as.data.frame(GetAssayData(object = juntti_OE_neuronal, slot = "data")) 
juntti_OE_neuronal_CRcounts <- juntti_OE_neuronal_CRs %>%
  filter(row.names(juntti_OE_neuronal_CRs) %in% c(V1R_genes$LOC, V2R_genes$Symbol, OR_genes$Symbol, TAAR_genes$Symbol)) %>%
  apply(1, function(x)sum(x!=0)) %>%
  as.data.frame()

colnames(juntti_OE_neuronal_CRcounts) <- c("counts")
juntti_OE_neuronal_CRcounts$LOC <- rownames(juntti_OE_neuronal_CRcounts)

ggplot(juntti_OE_neuronal_CRcounts, aes(x = 1:nrow(juntti_OE_neuronal_CRcounts), y = counts)) + 
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Individual CRs", y = "Number of Cells Per CR") + 
  theme_classic()

hist(juntti_OE_neuronal_CRcounts$counts, breaks = 100, main = "Histogram of Cells Per CR", xlab = "")
summary(juntti_OE_neuronal_CRcounts$counts)

```

Most CRs have a mean of ~48 neuronal cells which express them.



# What are the most represented ORs we have?

```{r}
#write.xlsx(juntti_OE_neuronal_CRcounts[order(juntti_OE_neuronal_CRcounts$counts, decreasing = TRUE), c("LOC", "counts")], file = "../excel/cellcounts_per_CR.xlsx")
ordered <- juntti_OE_neuronal_CRcounts[order(juntti_OE_neuronal_CRcounts$counts, decreasing = TRUE), c("LOC", "counts")]

kable(ordered[ordered$LOC %in% V1R_genes$LOC,], caption = "V1R Genes")
kable(ordered[ordered$LOC %in% V2R_genes$Symbol,], caption = "V2R Genes")
kable(ordered[ordered$LOC %in% OR_genes$Symbol,], caption = "OR Genes")
kable(ordered[ordered$LOC %in% TAAR_genes$Symbol,], caption = "TAAR Genes")
```




# Get average expression for each CR type

sum (or average) expression of each OR type, then plot on UMAP <br>

```{r}
features <- list("V1R_genes" = V1R_genes$LOC, "V2R_genes" = V2R_genes$Symbol, "OR_genes" = OR_genes$Symbol, "TAAR_genes" = TAAR_genes$Symbol)
juntti_OE_new <- genes_avg_expression(seur_obj = juntti_OE_new, features = features, slot = "data")
juntti_OE_neuronal2 <- subset(x = juntti_OE_new, idents = c(0,1))

FeaturePlot(juntti_OE_neuronal2, features = "V1R_genes_Sums")
FeaturePlot(juntti_OE_neuronal2, features = "V2R_genes_Sums")
FeaturePlot(juntti_OE_neuronal2, features = "OR_genes_Sums")
FeaturePlot(juntti_OE_neuronal2, features = "TAAR_genes_Sums")

FeaturePlot(juntti_OE_new, features = "V1R_genes_Sums")
FeaturePlot(juntti_OE_new, features = "V2R_genes_Sums")
FeaturePlot(juntti_OE_new, features = "OR_genes_Sums")
FeaturePlot(juntti_OE_new, features = "TAAR_genes_Sums")
```



# Save Seurat Object

```{r}
save(juntti_OE_new, file = "juntti_OR_interrogation_output.rda")
```


